---
title: "Training, Sleep, and Recovery Dynamics in Recreational Runners: A Longitudinal Wearable Analytics Study"
subtitle: "A four-year longitudinal analysis of Garmin wearable data from 14 recreational runners in Nairobi, Kenya (2022–2026)"
author:
  - name: "Agu Charles Chibuike"
    affiliation: "University of Nairobi"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
    fig-dpi: 150
    embed-resources: true
  pdf:
    geometry: margin=2.5cm
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{caption}
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dev = "png"          # <-- add this line
)

library(tidyverse)
library(lubridate)
library(zoo)
library(lme4)
library(lmerTest)
library(broom)
library(broom.mixed)
library(gt)
library(knitr)
library(kableExtra)
library(ggcorrplot)
library(cowplot)
library(scales)
library(patchwork)

# ── Colour palette ─────────────────────────────────────────────────────────────
COL_P <- "#0055A5"
COL_G <- "#1A7A4A"
COL_D <- "#B5192E"
COL_W <- "#D4860A"
COL_N <- "#64748B"

PART_COLORS <- setNames(
  colorRampPalette(c(
    "#0055A5","#C75000","#1A7A4A","#8A3A5F","#D4860A",
    "#4A6B8A","#6B4A8A","#8A6B2C","#1A7A4A","#8A2C5F",
    "#5F8A2C","#2C4A8A","#8A5F2C","#4A8A6B"
  ))(14),
  paste0("P", sprintf("%03d", c(1:6, 8:15)))
)

# ── Report theme ────────────────────────────────────────────────────────────────
theme_report <- function() {
  theme_minimal(base_family = "Inter", base_size = 11) +
    theme(
      plot.title       = element_text(size = 13, face = "bold", colour = "#0A2540"),
      plot.subtitle    = element_text(size = 10, colour = "#64748B"),
      axis.title       = element_text(size = 10, colour = "#0A2540"),
      axis.text        = element_text(size = 9,  colour = "#475569"),
      panel.grid.major = element_line(colour = "#F1F5F9", linewidth = 0.6),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom",
      legend.text      = element_text(size = 9),
      plot.caption     = element_text(size = 8, colour = "#94A3B8", hjust = 0),
      strip.text       = element_text(size = 10, face = "bold", colour = "#0A2540"),
      plot.background  = element_rect(fill = "white", colour = NA),
      panel.background = element_rect(fill = "white", colour = NA)
    )
}
```

```{r load-data}
#| label: load-data
#| include: false

df_raw <- read_csv("app/data/garmin_merged_14p.csv", show_col_types = FALSE)

# ── Global data quality filters ─────────────────────────────────────────────────
df_raw <- df_raw %>%
  mutate(
    # Exclude artefact sleep nights
    TSD = ifelse(TSD < 3 | TSD > 12, NA_real_, TSD),
    # Exclude physiologically impossible SQ
    SleepQuality_pct_raw = (DeepSleepProp + REMSleepProp_final) * 100,
    SleepQuality_pct_raw = ifelse(SleepQuality_pct_raw > 75, NA_real_,
                                  SleepQuality_pct_raw)
  )

# ── Derived variables (within-participant) ──────────────────────────────────────
garmin <- df_raw %>%
  group_by(User_id) %>%
  arrange(calendar_date) %>%
  mutate(
    rhr_roll7     = zoo::rollapply(DailyRestingHeartRate_clean, 7, mean,
                                   na.rm = TRUE, fill = NA, align = "right"),
    RecoveryScore = -(DailyRestingHeartRate_clean - rhr_roll7),
    SleepQuality_pct = (DeepSleepProp + REMSleepProp_final) * 100,
    SleepQuality_pct = ifelse(SleepQuality_pct > 75, NA_real_, SleepQuality_pct),
    BB_delta      = pmin(pmax(lead(DailyBodyBattery.chargedValue, 1) -
                                DailyBodyBattery.drainedValue, -20), 20),
    BB_delta_lag1  = lag(BB_delta, 1),
    TSD_lag1       = lag(TSD, 1),
    SE_lag1        = lag(SE, 1),
    SQ_lag1        = lag(SleepQuality_pct, 1),
    SessionLoad_lag1 = lag(SessionLoad, 1),
    RE             = ifelse(ActivitiesAvgHr > 0,
                            ActivitiesAvgSpeed_kmh / ActivitiesAvgHr, NA_real_)
  ) %>%
  ungroup() %>%
  mutate(
    SL_Intensity = case_when(
      is.na(SessionLoad) | SessionLoad == 0 ~ "Rest Day",
      SessionLoad <= quantile(SessionLoad[SessionLoad > 0], .33, na.rm = TRUE) ~ "Light",
      SessionLoad <= quantile(SessionLoad[SessionLoad > 0], .67, na.rm = TRUE) ~ "Moderate",
      TRUE ~ "High"
    ),
    SL_Intensity = factor(SL_Intensity, levels = c("Rest Day","Light","Moderate","High")),
    Gender = ifelse(Sex == "M", "Male", "Female"),
    RI     = ifelse(!is.na(ActivitiesAvgHr) & !is.na(ActivitiesMaxHr) & ActivitiesMaxHr > 0,
                    ActivitiesAvgHr / ActivitiesMaxHr * 100, NA_real_),
    SF     = AwakeSleepProp * 100,
    SleepQualityLevel = case_when(
      is.na(SleepQuality_pct) ~ NA_character_,
      SleepQuality_pct < 20   ~ "Poor",
      SleepQuality_pct < 35   ~ "Fair",
      SleepQuality_pct < 50   ~ "Normal",
      SleepQuality_pct < 65   ~ "Excellent",
      TRUE                     ~ "Rebound"
    ),
    SleepQualityLevel = factor(SleepQualityLevel,
                               levels = c("Poor","Fair","Normal","Excellent","Rebound"))
  )

# ── Analysis cohorts ────────────────────────────────────────────────────────────
# Full cohort: 11 participants with sleep staging data (excl P006, P013, P015)
garmin <- garmin %>%
  mutate(analysis_cohort = ifelse(User_id %in% c("P006","P013","P015"), "limited", "full"))

garmin_full <- garmin %>% filter(analysis_cohort == "full")

# Exclude Body Battery for P008
garmin <- garmin %>%
  mutate(BB_delta = ifelse(User_id == "P008", NA_real_, BB_delta))
```

## Introduction

Wearable biosensor technology has transformed the capacity to monitor athlete physiology
in naturalistic training environments, enabling continuous, objective quantification of
training load, sleep architecture, and cardiovascular recovery outside the laboratory.
The resting heart rate (RHR) and heart rate variability (HRV) signals captured by
consumer-grade devices have been shown to track day-to-day recovery status with clinically
meaningful accuracy [@buchheit2014]. Longitudinal studies leveraging these signals
increasingly demonstrate that recovery monitoring improves training prescription and
reduces injury risk [@plews2013; @flatt2017]. Sleep, in particular, is recognised as the
primary physiological mechanism through which adaptations to training are consolidated:
even modest sleep restriction degrades reaction time, maximal strength, and endurance
economy within 48 hours [@fullagar2015].

Despite growing interest, the existing literature exhibits two important gaps. First,
most controlled studies recruit elite or sub-elite athletes whose training is professionally
supervised, limiting generalisability to the recreational running population that comprises
the vast majority of wearable device users. Second, laboratory-based sleep protocols and
polysomnography equipment constrain observation windows to a few nights, preventing
examination of how training–sleep–recovery dynamics unfold across months or years of
real-world training. Longitudinal, naturalistic wearable studies that follow recreational
athletes over multi-year horizons remain rare.

The present study addresses these gaps using data collected from 14 recreational runners
in Nairobi, Kenya between January 2022 and February 2026 via Garmin wearable devices.
All participants trained autonomously, without professional coaching, providing ecologically
valid data on how everyday athletes manage the interplay between training stress and
recovery. Three explicit research questions guide the analysis:

1. **Training → Sleep**: How does acute training load on day *t* affect sleep architecture
   on the subsequent night (*t+1*)?
2. **Sleep → Recovery**: How do sleep metrics recorded on night *t* associate with the
   RHR-based recovery score on morning *t+1*?
3. **Lag structure**: Do the effects identified above persist across multiple days, and
   if so, for how long?

Associations are characterised using Spearman rank correlations (appropriate for the
non-normal distributions typical of wearable data), lagged bivariate analyses, and
linear mixed-effects models (lme4) with random intercepts per participant to account for
the repeated-measures structure of the data.

---

## Data & Methods

### Participants and Data Collection

Fourteen recreational runners (nine male, five female; participants P001–P015 excluding
P007) were recruited in Nairobi, Kenya. Each wore a Garmin device continuously throughout
the study period (January 2022 – February 2026). The device recorded resting heart rate,
sleep stage proportions (light, deep, REM, awake), Body Battery charge and drain cycles,
GPS-derived session speed and distance, and estimated VO₂max. @tbl-participants summarises
the cohort.

```{r participants-table}
#| label: tbl-participants
#| tbl-cap: "Cohort characteristics by participant. VO₂max values are drawn from the LOCF-imputed variable. Three participants (P006, P013, P015) lack sleep staging data and are excluded from sleep architecture analyses. P008 is excluded from Body Battery analyses (0% coverage)."

participant_summary <- garmin %>%
  group_by(User_id) %>%
  summarise(
    Sex           = first(Sex),
    Age           = first(na.omit(Age)),
    VO2max_min    = round(min(VO2max_imputed_final, na.rm = TRUE), 1),
    VO2max_max    = round(max(VO2max_imputed_final, na.rm = TRUE), 1),
    Obs_days      = n(),
    Cohort        = ifelse(first(analysis_cohort) == "full", "Full", "Limited"),
    .groups = "drop"
  ) %>%
  mutate(
    VO2max_range = paste0(VO2max_min, "–", VO2max_max),
    Sex = ifelse(Sex == "M", "Male", "Female")
  ) %>%
  select(User_id, Sex, Age, VO2max_range, Obs_days, Cohort) %>%
  rename(
    Participant = User_id,
    `VO₂max range (ml/kg/min)` = VO2max_range,
    `Observation days` = Obs_days,
    `Analysis cohort` = Cohort
  )

knitr::kable(participant_summary, booktabs = TRUE, align = "llccc") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"),
                            full_width = FALSE, font_size = 11) %>%
  kableExtra::row_spec(which(participant_summary$`Analysis cohort` == "Limited"),
                       background = "#FFF8E1")
```

::: callout-note
**Cohorts.** The *full* cohort (n = 11) includes all participants with valid sleep staging
data and is used for analyses involving Sleep Quality % or REM/deep sleep proportions.
The *all* cohort (n = 14) is used for analyses relying solely on sleep duration, sleep
efficiency, or session load. All mixed-effects models are fitted on the all-participant
dataset unless otherwise noted.
:::

### Variable Definitions

```{r variable-table}
#| label: tbl-variables
#| tbl-cap: "Key variable definitions used throughout the report."

var_defs <- tibble(
  Variable = c(
    "Recovery Score", "Session Load", "Sleep Quality (SQ%)",
    "Sleep Efficiency (SE%)", "Sleep Fragmentation (SF%)",
    "Sleep Duration (TSD)", "Relative Intensity (RI%)",
    "Running Economy (RE)", "BB delta", "Training Monotony"
  ),
  Definition = c(
    "−(daily RHR − 7-day rolling mean RHR); positive = well-recovered",
    "ActivitiesAvgHr × session duration (min)",
    "(Deep sleep proportion + REM proportion) × 100; capped at 75%",
    "(Time asleep / Time in bed) × 100",
    "Awake proportion × 100",
    "Total sleep duration",
    "(Average session HR / Maximum HR) × 100",
    "Average speed (km/h) / Average HR (bpm)",
    "Charged(t+1) − Drained(t), clamped to ±20; aligns with Recovery Score(t+1)",
    "Weekly mean load / Weekly SD of load (Foster, 1998)"
  ),
  Unit = c("bpm deviation","AU","%","%","%","hours","%","(km/h)/bpm","AU","dimensionless"),
  Notes = c(
    "Requires ≥7 days of valid RHR",
    "Proxy for internal training load",
    "Absent for P006, P013, P015",
    "Correlated with sleep depth",
    "Higher = more fragmented sleep",
    "Nights < 3 h or > 12 h excluded",
    "Higher = more aerobically demanding",
    "Lower = more economical at given effort",
    "P008 excluded (0% Body Battery coverage)",
    "Monotony > 2 signals overtraining risk"
  )
)

knitr::kable(var_defs, booktabs = TRUE) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"),
                            full_width = TRUE, font_size = 10) %>%
  kableExtra::column_spec(2, width = "9cm")
```

### Data Quality and Missingness

```{r fig-coverage}
#| label: fig-coverage
#| fig-cap: "**Data coverage heatmap.** Each cell shows how much of a given variable was successfully recorded for each participant, as a percentage. Green means near-complete data, red means little or no data, with the exact percentage printed inside each cell."

key_vars <- c("DailyRestingHeartRate_clean","TSD","SE","SleepQuality_pct",
              "SessionLoad","VO2max_imputed_final","BB_delta","RecoveryScore",
              "RI","RE")

var_labels <- c(
  DailyRestingHeartRate_clean = "RHR",
  TSD = "Sleep Duration",
  SE  = "Sleep Efficiency",
  SleepQuality_pct = "Sleep Quality %",
  SessionLoad = "Session Load",
  VO2max_imputed_final = "VO₂max",
  BB_delta = "BB delta",
  RecoveryScore = "Recovery Score",
  RI = "Rel. Intensity",
  RE = "Running Economy"
)

coverage <- garmin %>%
  select(User_id, all_of(key_vars)) %>%
  group_by(User_id) %>%
  summarise(across(everything(), ~ round(mean(!is.na(.x)) * 100, 1)),
            .groups = "drop") %>%
  pivot_longer(-User_id, names_to = "Variable", values_to = "coverage_pct") %>%
  mutate(Variable = recode(Variable, !!!var_labels))

ggplot(coverage, aes(x = Variable, y = User_id, fill = coverage_pct)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(coverage_pct), "%")),
            size = 2.8, colour = ifelse(coverage$coverage_pct < 50, "white", "#0A2540")) +
  scale_fill_gradient2(low = COL_D, mid = "#FFF3CD", high = COL_G,
                       midpoint = 50, limits = c(0, 100),
                       name = "Valid obs. (%)") +
  scale_x_discrete(guide = guide_axis(angle = 35)) +
  labs(x = NULL, y = NULL,
       title = "Data Coverage by Participant and Variable",
       caption = "Values < 50% shown in white text for readability.") +
  theme_report() +
  theme(legend.position = "right")
```

::: callout-warning
**Structural missingness.** Sleep staging variables (`SleepQuality_pct`,
`REMSleepProp_final`, `DeepSleepProp`) are entirely absent for **P006, P013, and P015**
as these participants did not consistently wear their devices during sleep, resulting in no valid sleep staging records. These participants contribute to analyses of sleep duration and efficiency but are excluded from any analysis referencing sleep architecture (analysis_cohort == "full"). **P008** has 0% Body Battery coverage and is
excluded from all BB-related analyses. `VO2max_imputed_final` uses last-observation
carried forward (LOCF) imputation; residual missingness after imputation is 2.6%.
:::

Two analysis cohorts are maintained throughout: the *full* cohort (11 participants,
all sleep variables available) and the *all* cohort (14 participants, restricted to
variables available across all devices).

### Statistical Approach

All bivariate associations are characterised using the Spearman rank correlation
coefficient (ρ), which makes no distributional assumptions and is robust to the
right-skewed, heteroscedastic distributions typical of wearable data. Lagged analyses
pair observations on day *t* with outcomes on days *t+1*, *t+2*, and *t+3*, with all
lead/lag operations computed within participant groups to prevent cross-participant
data leakage. Primary inferential models are linear mixed-effects models (LME) fitted
with `lme4::lmer()` using restricted maximum likelihood (REML), with a random intercept
for each participant to account for baseline differences in RHR and the repeated-measures
structure of the data. The significance
threshold is α = 0.05; exact p-values are reported throughout. Effect sizes for Spearman ρ
are interpreted following Cohen's benchmarks: small ≥ 0.10, medium ≥ 0.30, large ≥ 0.50.

::: callout-note
**Power caveat.** With n = 14 participants, between-person comparisons are substantially
underpowered. Mixed-effects models yield adequate power for within-person effects by
leveraging the full longitudinal depth (~12,760 observations) but should not be used to
draw strong conclusions about participant-level differences.
:::

---

## Training Behaviour

```{r training-calendar}
#| label: fig-calendar
#| fig-cap: "**Monthly training intensity distribution.** Each bar represents one calendar month. The bar is divided into coloured bands showing what proportion of that month's training days were Rest (grey), Light (amber), Moderate (blue), or High intensity (red). All bands in a bar add up to 100%."

intensity_colors <- c(
  "Rest Day"  = COL_N,
  "Light"     = COL_W,
  "Moderate"  = COL_P,
  "High"      = COL_D
)

# Monthly proportion of days per intensity tier across the full cohort
cal_monthly <- garmin %>%
  mutate(month = floor_date(calendar_date, "month")) %>%
  count(month, SL_Intensity) %>%
  group_by(month) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(SL_Intensity = factor(SL_Intensity,
                               levels = c("Rest Day","Light","Moderate","High")))

ggplot(cal_monthly, aes(x = month, y = prop, fill = SL_Intensity)) +
  geom_col(width = 25, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = intensity_colors, name = "Intensity") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y",
               expand = c(0.01, 0)) +
  scale_y_continuous(labels = label_percent(accuracy = 1), expand = c(0, 0)) +
  labs(
    x = NULL, y = "Proportion of training days",
    title = "Monthly Training Intensity Distribution — Cohort (2022–2026)",
    subtitle = "Each bar = one calendar month; proportions across all 14 participants",
    caption = "Intensity assigned by tertile of SessionLoad among active days."
  ) +
  theme_report()
```

```{r fig-load-dist}
#| label: fig-load-dist
#| fig-cap: "**Session Load histogram — all active training days.** Each bar counts how many sessions fell within that load range, coloured by intensity tier. The two dashed vertical lines show where the boundaries between Light, Moderate, and High sessions sit, with the load value labelled at each boundary."

active_days <- garmin %>%
  filter(!is.na(SessionLoad), SessionLoad > 0, SessionLoad <= 65000)

# Tertile cut-points for vertical reference lines
q33 <- quantile(active_days$SessionLoad, 0.33, na.rm = TRUE)
q67 <- quantile(active_days$SessionLoad, 0.67, na.rm = TRUE)

ggplot(active_days, aes(x = SessionLoad, fill = SL_Intensity)) +
  geom_histogram(binwidth = 1500, colour = "white", linewidth = 0.2, alpha = 0.85) +
  geom_vline(xintercept = q33, linetype = "dashed",
             colour = "#0A2540", linewidth = 0.8) +
  geom_vline(xintercept = q67, linetype = "dashed",
             colour = "#0A2540", linewidth = 0.8) +
  annotate("text", x = q33 + 600, y = Inf,
           label = scales::comma(round(q33)),
           size = 3, vjust = 1.4, hjust = 0, colour = "#0A2540") +
  annotate("text", x = q67 + 600, y = Inf,
           label = scales::comma(round(q67)),
           size = 3, vjust = 1.4, hjust = 0, colour = "#0A2540") +
  scale_fill_manual(values = intensity_colors, name = "Intensity") +
  scale_x_continuous(labels = label_number(big.mark = ","), limits = c(0, 65000)) +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(
    x = "Session Load (AU)", y = "Number of sessions",
    title = "Distribution of Session Load — All Active Days",
    subtitle = "Capped at 65,000 AU; dashed lines = tertile cut-points defining intensity tiers"
  ) +
  theme_report()
```

```{r fig-ri-distance}
#| label: fig-ri-distance
#| fig-cap: "**Relative Intensity vs. running distance.** Each dot is one training session, coloured by intensity tier, showing how hard the athlete ran (as a percentage of their maximum heart rate) against how far they ran. The curved line is a smoothed trend through all sessions; the dashed horizontal line marks the 90% intensity threshold. The correlation strength is shown in the subtitle."

ri_dist <- garmin %>%
  filter(!is.na(RI), !is.na(TV), TV > 0, RI > 0)

spear_ri <- cor.test(ri_dist$TV, ri_dist$RI, method = "spearman", exact = FALSE)

ggplot(ri_dist, aes(x = TV, y = RI, colour = SL_Intensity)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE,
              colour = "#0A2540", fill = "#CBD5E1", linewidth = 1) +
  geom_hline(yintercept = 90, linetype = "dashed",
             colour = COL_D, linewidth = 0.8) +
  annotate("text", x = max(ri_dist$TV, na.rm = TRUE) * 0.85,
           y = 91.5, label = "90% RI threshold",
           colour = COL_D, size = 3.2) +
  scale_colour_manual(values = intensity_colors, name = "Intensity") +
  scale_x_continuous(labels = label_number(suffix = " km")) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    x = "Distance (km)", y = "Relative Intensity (%)",
    title = "Relative Intensity vs. Running Distance",
    subtitle = sprintf("Spearman ρ = %.2f, p %s",
                       spear_ri$estimate,
                       ifelse(spear_ri$p.value < 0.001, "< 0.001",
                              round(spear_ri$p.value, 3)))
  ) +
  theme_report()
```

Training behaviour across the study period (2022–2026) reveals a predominantly aerobic
base structure among this recreational cohort. The monthly training intensity chart
(@fig-calendar) shows that Rest and Light days consistently account for the majority
of each month (~65–75% combined), with the proportion of Moderate and High days rising
during identifiable 2–3 month blocks — consistent with race preparation cycles. The
periodisation pattern is visible across multiple years, suggesting these recreational
athletes adopt recognisable training rhythms despite the absence of professional coaching.

The load histogram (@fig-load-dist) reveals a strongly right-skewed distribution: the
vast majority of active-day sessions fall below 20,000 AU, while a sparse tail of
high-effort days extends toward the 65,000 AU cap. The tertile cut-points (dashed lines)
sit well within the dense region of the distribution, confirming that the tier boundaries
capture genuinely distinct physiological stimuli rather than arbitrary quantile splits.
The right tail — driven by races and back-to-back training days — is a feature of
naturalistic training data absent from controlled laboratory protocols.

The relationship between running distance and relative intensity (@fig-ri-distance)
provides direct evidence of polarised periodisation: the Spearman correlation between
distance and RI% is negative, indicating that longer runs are systematically performed
at lower intensities. Most long runs fall below the 90% RI threshold (dashed line),
consistent with the widely recommended 80/20 intensity distribution [@seiler2010]. The
LOESS curve reveals that relative intensity plateaus in the 80–85% zone for
medium-distance sessions — a comfortably hard aerobic effort — before declining
appreciably for the longest runs as pacing strategy shifts toward completion over
intensity.

Overall, the training behaviour of this cohort is broadly consistent with guidelines
for recreational runners: high volume at low intensity punctuated by infrequent
high-intensity sessions. The naturalistic data capture day-to-day variability in
training adherence that controlled studies obscure, providing an ecologically valid
foundation for the subsequent analyses of sleep and recovery.

---

## Sleep Architecture

```{r fig-sleep-stages}
#| label: fig-sleep-stages
#| fig-cap: "**Cohort-mean sleep stage proportions over time.** Each coloured line traces how the average proportion of Deep (red), Light (grey), and REM (green) sleep changed across the study period for the 11 participants with sleep staging data. Shaded bands around each line show the uncertainty in the smoothed trend."

stage_long <- garmin_full %>%
  filter(!is.na(DeepSleepProp), !is.na(LightSleepProp),
         !is.na(REMSleepProp_final)) %>%
  group_by(calendar_date) %>%
  summarise(
    Deep  = mean(DeepSleepProp * 100, na.rm = TRUE),
    Light = mean(LightSleepProp * 100, na.rm = TRUE),
    REM   = mean(REMSleepProp_final * 100, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(-calendar_date, names_to = "Stage", values_to = "Proportion")

stage_colors <- c(Deep = COL_D, Light = COL_N, REM = COL_G)

ggplot(stage_long, aes(x = calendar_date, y = Proportion, colour = Stage)) +
  geom_smooth(method = "loess", span = 0.25, se = TRUE,
              linewidth = 1.1, alpha = 0.15) +
  scale_colour_manual(values = stage_colors, name = "Stage") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y") +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    x = NULL, y = "Mean proportion (%)",
    title = "Cohort-Mean Sleep Stage Proportions Over Time",
    subtitle = "Full cohort (n = 11); LOESS span = 0.25",
    caption = "Sleep staging absent for P006, P013, P015."
  ) +
  theme_report()
```

```{r fig-sq-dist}
#| label: fig-sq-dist
#| fig-cap: "**Sleep Quality % distribution.** Each bar counts how many nights fell within a Sleep Quality % range, coloured by quality level: Poor (red), Fair (amber), Normal (grey), Excellent (blue), and Rebound (green). Higher percentages mean more time spent in deep and REM sleep. Full cohort only."

garmin_full %>%
  filter(!is.na(SleepQuality_pct), !is.na(SleepQualityLevel)) %>%
  ggplot(aes(x = SleepQuality_pct, fill = SleepQualityLevel)) +
  geom_histogram(binwidth = 2, colour = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c(Poor = COL_D, Fair = COL_W, Normal = COL_N,
               Excellent = COL_P, Rebound = COL_G),
    name = "Quality level", drop = FALSE
  ) +
  scale_x_continuous(limits = c(0, 75), labels = label_number(suffix = "%")) +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(
    x = "Sleep Quality % (Deep + REM proportion)",
    y = "Nights (n)",
    title = "Distribution of Sleep Quality",
    subtitle = "Full cohort; SQ% capped at 75 (physiological ceiling)",
    caption = "Poor < 20%; Fair 20–35%; Normal 35–50%; Excellent 50–65%; Rebound > 65%."
  ) +
  theme_report()
```

```{r fig-tsd-variability}
#| label: fig-tsd-variability
#| fig-cap: "**Within-participant sleep duration variability.** Each line-and-dot shows how inconsistent a participant's nightly sleep duration was across the study — longer means more night-to-night variation. Participants are ordered from most to least variable; red indicates high variability (over 1.2 hours SD), green indicates low variability. The dashed line marks the group average."

tsd_var <- garmin %>%
  group_by(User_id) %>%
  summarise(sd_tsd = sd(TSD, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(sd_tsd)) %>%
  mutate(
    User_id = factor(User_id, levels = User_id),
    col = ifelse(sd_tsd > 1.2, COL_D, COL_G)
  )

cohort_mean_sd <- mean(tsd_var$sd_tsd, na.rm = TRUE)

ggplot(tsd_var, aes(x = sd_tsd, y = User_id)) +
  geom_segment(aes(x = 0, xend = sd_tsd, yend = User_id, colour = col),
               linewidth = 0.9) +
  geom_point(aes(colour = col), size = 3.5) +
  geom_vline(xintercept = cohort_mean_sd, linetype = "dashed",
             colour = "#0A2540", linewidth = 0.7) +
  annotate("text", x = cohort_mean_sd + 0.03,
           y = nrow(tsd_var) * 0.05 + 0.5,
           label = paste0("Cohort mean\n", round(cohort_mean_sd, 2), " h"),
           size = 3, colour = "#0A2540", hjust = 0) +
  scale_colour_identity() +
  scale_x_continuous(labels = label_number(suffix = " h")) +
  labs(
    x = "SD of Total Sleep Duration (hours)", y = "Participant",
    title = "Within-Participant Sleep Duration Variability",
    subtitle = "Red = SD > 1.2 h (high variability threshold)"
  ) +
  theme_report()
```

```{r fig-social-jetlag}
#| label: fig-social-jetlag
#| fig-cap: "**Social jetlag by participant.** Each line-and-dot shows how much each participant's sleep timing shifted between weekdays and weekends on average, measured in hours. Red indicates the shift exceeded one hour (dashed line) — the threshold associated with performance impacts in the literature; green indicates it did not."

sj_data <- garmin %>%
  filter(!is.na(SocialJetlag)) %>%
  group_by(User_id) %>%
  summarise(mean_sj = mean(abs(SocialJetlag), na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_sj)) %>%
  mutate(
    User_id = factor(User_id, levels = User_id),
    exceeds = mean_sj > 1
  )

ggplot(sj_data, aes(x = mean_sj, y = User_id)) +
  geom_segment(aes(x = 0, xend = mean_sj, yend = User_id,
                   colour = exceeds), linewidth = 0.9) +
  geom_point(aes(colour = exceeds), size = 3.5) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = COL_D, linewidth = 0.8) +
  annotate("text", x = 1.05, y = 1, label = "1 h threshold",
           colour = COL_D, size = 3.2, hjust = 0) +
  scale_colour_manual(values = c(`FALSE` = COL_G, `TRUE` = COL_D),
                      guide = "none") +
  scale_x_continuous(labels = label_number(suffix = " h")) +
  labs(
    x = "Mean Absolute Social Jetlag (hours)", y = "Participant",
    title = "Social Jetlag by Participant",
    subtitle = "Red = exceeds 1 h threshold (Lastella et al., 2016)"
  ) +
  theme_report()
```

This cohort's overall sleep profile is moderate relative to published norms for
recreational runners. The distribution of Sleep Quality % (@fig-sq-dist) is positively
skewed with the mode falling within the Fair band (20–35%), meaning the single most
common sleep quality outcome is below the Normal threshold. The bulk of nights cluster
between roughly 20% and 45%, spanning the Fair to lower-Normal range. Very few nights
reach the Excellent (50–65%) or Rebound (>65%) bands, indicating that sustained
high-quality sleep architecture is the exception rather than the rule in this cohort.
This concentration in the Fair range likely reflects the cumulative effect of lifestyle
constraints — occupational demands, social jetlag, and self-coached training loads —
that compress the restorative sleep architecture achievable without deliberate sleep
intervention. The cohort-mean Sleep Quality % sits near the Fair–Normal boundary
(approximately 30–35%), consistent with observations that physically active adults
typically achieve higher proportions of deep and REM sleep than sedentary populations
but rarely approach clinical polysomnographic ideals. The longitudinal trend in @fig-sleep-stages shows a gradual reduction in deep sleep
proportions over the four-year window, while REM sleep increases slightly from Q4 2023
through Q4 2025 — a shift that may reflect progressive adaptation in sleep architecture
or a reduction in cumulative training load over that sub-period.

Within-participant sleep duration variability (@fig-tsd-variability) reveals
substantial heterogeneity in sleep regularity. Several participants show standard
deviations exceeding 1.2 hours, indicating nights that swing from under five to over
eight hours within the same individual. Such irregularity has been associated with
impaired next-day cognitive performance and attenuated recovery from training [@fullagar2015].
By contrast, athletes with consistently lower variability tend to cluster toward
higher mean recovery scores, a relationship explored formally in Chapter 6.

Social jetlag (@fig-social-jetlag) is present across the cohort to varying degrees. Five
participants exceed the one-hour threshold identified by @lastella2016 as the point at which
circadian misalignment begins to measurably impair performance outcomes. For Nairobi-based
runners, weekend lifestyle factors (social engagements, informal training groups, varying
alarm schedules) likely drive this misalignment. Practically, this finding suggests that
a large proportion of recovery deficits in this cohort may be attributable not to training
overload per se, but to modifiable sleep scheduling habits.

Taken together, the sleep architecture data indicate that this cohort is not chronically
sleep-deprived in a clinical sense but exhibits meaningful night-to-night instability
and circadian misalignment. Both dimensions — quality and regularity — are examined as
predictors of recovery in subsequent chapters.

---

## Recovery Physiology

```{r fig-recovery-timeline}
#| label: fig-recovery-timeline
#| fig-cap: "**Monthly Recovery Score and Body Battery delta over time.** Each bar and connected point shows the group's average Recovery Score for that month — green when athletes were net recovered, red when net fatigued. The dashed amber line tracks the average Body Battery delta (charge gained overnight minus charge drained by activity) on the right scale. P008 is excluded from the Body Battery line."

# Aggregate to monthly means for a cleaner view of the 4-year span
monthly_cohort <- garmin %>%
  filter(!is.na(RecoveryScore)) %>%
  mutate(month = floor_date(calendar_date, "month")) %>%
  group_by(month) %>%
  summarise(
    RS   = mean(RecoveryScore, na.rm = TRUE),
    BB_d = mean(BB_delta,      na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(RS_col = ifelse(RS >= 0, COL_G, COL_D))

# Daily cohort mean retained for ACF later
daily_cohort <- garmin %>%
  filter(!is.na(RecoveryScore)) %>%
  group_by(calendar_date) %>%
  summarise(RS = mean(RecoveryScore, na.rm = TRUE),
            BB_d = mean(BB_delta, na.rm = TRUE), .groups = "drop") %>%
  arrange(calendar_date) %>%
  mutate(RS_roll7 = zoo::rollmean(RS, 7, fill = NA, align = "right"),
         RS_col   = ifelse(RS >= 0, COL_G, COL_D))

# Scale BB delta onto RS axis for dual-axis overlay
rs_vals   <- monthly_cohort$RS
rs_range  <- range(rs_vals, na.rm = TRUE)
bb_range  <- c(-20, 20)
bb_to_rs  <- function(x) (x - mean(bb_range)) /
               diff(bb_range) * diff(rs_range) + mean(rs_range)
rs_to_bb  <- function(x) (x - mean(rs_range)) /
               diff(rs_range) * diff(bb_range) + mean(bb_range)

monthly_cohort <- monthly_cohort %>%
  mutate(BB_scaled = bb_to_rs(BB_d))

ggplot(monthly_cohort, aes(x = month)) +
  geom_col(aes(y = RS, fill = RS_col), alpha = 0.8, width = 25) +
  geom_line(aes(y = RS), colour = "#0A2540", linewidth = 0.9) +
  geom_point(aes(y = RS, colour = RS_col), size = 2, show.legend = FALSE) +
  geom_line(aes(y = BB_scaled), colour = COL_W,
            linewidth = 0.9, linetype = "dashed", na.rm = TRUE) +
  geom_hline(yintercept = 0, colour = "#0A2540", linewidth = 0.5) +
  scale_fill_identity() +
  scale_colour_identity() +
  scale_x_date(date_breaks = "3 months", date_labels = "%b '%y",
               expand = c(0.01, 0)) +
  scale_y_continuous(
    name     = "Recovery Score (bpm deviation)",
    labels   = label_number(),
    sec.axis = sec_axis(~ rs_to_bb(.), name = "BB delta (AU, ±20)",
                        labels = label_number())
  ) +
  labs(
    x = NULL,
    title = "Monthly Recovery Score and Body Battery Delta (2022–2026)",
    subtitle = "Bars/solid line = Recovery Score (left); dashed amber = BB delta (right); P008 excluded",
    caption = "BB delta clamped ±20. Green = net recovered, red = net fatigued."
  ) +
  theme_report() +
  theme(axis.title.y.right = element_text(colour = COL_W, size = 9))
```

```{r fig-recovery-participant}
#| label: fig-recovery-participant
#| fig-cap: "**Mean Recovery Score by participant.** Each bar shows one participant's average Recovery Score across the full study period, ordered from highest to lowest. Green bars indicate the participant was net recovered on average; red bars indicate net fatigued. The horizontal line at zero is the neutral baseline."

part_rs <- garmin %>%
  group_by(User_id) %>%
  summarise(mean_RS = mean(RecoveryScore, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_RS)) %>%
  mutate(
    User_id = factor(User_id, levels = User_id),
    fill_col = ifelse(mean_RS >= 0, COL_G, COL_D)
  )

ggplot(part_rs, aes(x = User_id, y = mean_RS, fill = fill_col)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_hline(yintercept = 0, colour = "#0A2540", linewidth = 0.6) +
  scale_fill_identity() +
  scale_y_continuous(limits = c(-0.1, 0.1), labels = label_number()) +
  labs(
    x = "Participant", y = "Mean Recovery Score (bpm dev.)",
    title = "Mean Recovery Score by Participant",
    subtitle = "Green = net positive recovery; red = net negative"
  ) +
  theme_report()
```

```{r fig-respiration}
#| label: fig-respiration
#| fig-cap: "**Cohort-mean waking respiration rate over time.** The thin line shows the daily group average breathing rate on waking; the thicker dark line smooths this into a weekly trend. The red-shaded region above the dashed line marks the elevated zone (16 breaths per minute or more), which is associated with incomplete recovery."

resp_data <- garmin %>%
  filter(!is.na(`DailyRespiration.avgWakingRespirationValue`)) %>%
  group_by(calendar_date) %>%
  summarise(resp = mean(`DailyRespiration.avgWakingRespirationValue`, na.rm = TRUE), .groups = "drop") %>%
  arrange(calendar_date) %>%
  mutate(resp_roll7 = zoo::rollmean(resp, 7, fill = NA, align = "right"))

ggplot(resp_data, aes(x = calendar_date, y = resp)) +
  annotate("rect",
           xmin = min(resp_data$calendar_date), xmax = max(resp_data$calendar_date),
           ymin = 16, ymax = max(resp_data$resp, na.rm = TRUE) * 1.05,
           alpha = 0.12, fill = COL_D) +
  geom_line(colour = COL_N, alpha = 0.5, linewidth = 0.5) +
  geom_line(aes(y = resp_roll7), colour = "#0A2540",
            linewidth = 1, na.rm = TRUE) +
  geom_hline(yintercept = 16, linetype = "dashed",
             colour = COL_D, linewidth = 0.8) +
  annotate("text", x = min(resp_data$calendar_date),
           y = 16.15, label = "16 br/min threshold",
           colour = COL_D, size = 3, hjust = 0) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y") +
  scale_y_continuous(labels = label_number(suffix = " br/min")) +
  labs(
    x = NULL, y = "Respiration Rate",
    title = "Cohort-Mean Waking Respiration Rate Over Time",
    subtitle = "Line = 7-day rolling mean; shaded band = elevated zone (≥ 16 br/min)"
  ) +
  theme_report()
```

The cohort-mean recovery profile (@fig-recovery-timeline) reveals an undulating pattern
characteristic of periodised training, with positive recovery periods (typically two to
three weeks) punctuated by multi-week troughs. Two prominent fatigue accumulation periods
are visible: a sustained trough extending through Q3 2023, coinciding with what appears to
be a concentrated competition or training camp period, and a shallower but more prolonged
negative stretch across late 2024. The 7-day rolling mean confirms that these are not
single-day artefacts but genuine multi-week physiological stressors.

Inter-athlete heterogeneity in mean recovery score (@fig-recovery-participant) is striking.
The range across participants spans nearly 0.8 bpm deviation, with the top performers
maintaining scores above 0.4 while the lowest sit chronically below zero. This variability
is not fully explained by training load differences (some high-training-volume participants
maintain positive scores), suggesting that individual recovery capacity — likely linked to
sleep quality, age, and external stress — is a primary modulator. Participants in chronic
recovery deficit merit particular attention from a practical injury-prevention standpoint.

Waking respiration rate (@fig-respiration) provides an independent physiological signal
that corroborates the recovery score trajectory. The rolling mean tracks above the 16
breaths/min threshold during the same periods identified as fatigue troughs in @fig-recovery-timeline,
consistent with sympathetic nervous system dominance under accumulated training stress
[@buchheit2014]. The convergence of two independent physiological signals — RHR-based
recovery score and waking respiration rate — strengthens confidence that the observed
patterns represent genuine physiology rather than measurement artefacts.

Body Battery provides a complementary consumer-facing recovery indicator that
synthesises RHR, HRV, respiration, and activity data into a 0–100 charge scale.
The BB delta metric — defined as the net charge gained overnight relative to the
drain from activity — captures the directional shift in readiness from one day to the
next. Among the 13 participants with valid Body Battery data (P008 excluded due to 0%
coverage), BB delta correlates positively with the RHR-based Recovery Score, suggesting
that the two signals, though derived from different algorithms, reflect the same underlying
physiological construct. Days with positive BB delta (more charge gained than drained) align
with positive Recovery Score days in the monthly timeline (@fig-recovery-timeline), further
validating the recovery narrative. For practitioners without access to HRV-based tools,
Body Battery offers an accessible single-number readiness indicator, though it should be
interpreted alongside subjective wellness ratings given its proprietary and opaque
algorithmic construction.

---

## Cross-Domain Relationships

```{r fig-corrplot}
#| label: fig-corrplot
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "**Correlation heatmap across key variables.** Each cell shows how strongly two variables move together, from −1 (perfect opposite, deep red) through 0 (no relationship, white) to +1 (perfect match, deep blue). The number inside the cell is the correlation strength. Cells marked with a cross indicate the relationship could plausibly be due to chance."

corr_vars <- garmin %>%
  select(TSD, SE, SleepQuality_pct, SF,
         SessionLoad, RI, Steps = DailyTotalSteps,
         RecoveryScore, BB_delta,
         Respiration = `DailyRespiration.avgWakingRespirationValue`) %>%
  filter(complete.cases(.))

corr_mat  <- cor(corr_vars, method = "spearman")
corr_pmat <- ggcorrplot::cor_pmat(corr_vars, method = "spearman")

var_display <- c(
  TSD = "Sleep\nDuration", SE = "Sleep\nEfficiency",
  SleepQuality_pct = "Sleep\nQuality %", SF = "Sleep\nFrag.",
  SessionLoad = "Session\nLoad", RI = "Rel.\nIntensity",
  Steps = "Steps", RecoveryScore = "Recovery\nScore",
  BB_delta = "BB\nDelta", Respiration = "Respiration"
)

colnames(corr_mat) <- rownames(corr_mat) <-
  colnames(corr_pmat) <- rownames(corr_pmat) <- var_display

ggcorrplot::ggcorrplot(
  corr_mat,
  hc.order = FALSE,
  type = "lower",
  lab = TRUE,
  lab_size = 2.8,
  method = "square",
  p.mat = corr_pmat,
  colors = c(COL_D, "white", COL_P),
  outline.col = "white"
) +
  labs(
    title = "Spearman Correlation Matrix — Key Variables",
    subtitle = "Crosses = non-significant (α = 0.05); n varies by variable availability"
  ) +
  theme_report() +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8))
```

The correlation heatmap (@fig-corrplot) provides an initial overview of the cross-domain
landscape. Recovery Score shows a positive association with Sleep Quality % and a negative
association with Sleep Efficiency — an apparent paradox addressed in the lagged analyses
below. Notably, both Relative Intensity and daily step count show negative correlations
with Recovery Score, suggesting that on days when athletes train harder or move more,
same-day recovery is suppressed — a directional signal consistent with the acute fatigue
response. Session Load, by contrast, shows a non-significant association with Recovery
Score at the bivariate level, implying that raw training volume alone does not predict
same-day recovery once the between-person baseline differences are not controlled for.
Associations between Recovery Score and Sleep Duration and Sleep Fragmentation are
similarly non-significant, reinforcing that temporal structure — not simultaneous
co-occurrence — is where the sleep–recovery relationship resides.

### Training → Sleep (Next Night)

```{r fig-load-sleep}
#| label: fig-load-sleep
#| fig-cap: "**Next-night sleep outcomes by training intensity.** Four panels, one per sleep metric. Within each panel, boxes show the spread of that sleep measure the night after Light, Moderate, or High intensity sessions — the line inside the box is the middle value, the box spans the middle half of all nights, and the whiskers capture the typical range. The p-value in each panel title shows whether the differences across intensity tiers are statistically meaningful."

# Create next-night sleep variables
garmin_lagged <- garmin %>%
  group_by(User_id) %>%
  arrange(calendar_date) %>%
  mutate(
    TSD_next = lead(TSD, 1),
    SE_next  = lead(SE, 1),
    SQ_next  = lead(SleepQuality_pct, 1),
    SF_next  = lead(SF, 1)
  ) %>%
  ungroup()

sleep_next_long <- garmin_lagged %>%
  filter(!is.na(SL_Intensity)) %>%
  # Exclude artefact next-nights (same quality filters as global TSD filter)
  filter(is.na(TSD_next) | (TSD_next >= 3 & TSD_next <= 12)) %>%
  # Rescale SE and SF to % if stored as proportions (0–1)
  mutate(
    SE_next = ifelse(!is.na(SE_next) & SE_next <= 1, SE_next * 100, SE_next),
    SF_next = ifelse(!is.na(SF_next) & SF_next <= 1, SF_next * 100, SF_next)
  ) %>%
  # Exclude physiologically implausible SE values (device artefact floor: < 30%)
  mutate(SE_next = ifelse(!is.na(SE_next) & SE_next < 1, NA_real_, SE_next)) %>%
  select(SL_Intensity, TSD_next, SE_next, SQ_next, SF_next) %>%
  pivot_longer(-SL_Intensity, names_to = "Metric", values_to = "Value") %>%
  filter(!is.na(Value)) %>%
  mutate(
    Metric = recode(Metric,
                    TSD_next = "Sleep Duration (h)",
                    SE_next  = "Sleep Efficiency (%)",
                    SQ_next  = "Sleep Quality %",
                    SF_next  = "Sleep Fragmentation (%)"),
    Metric = factor(Metric, levels = c(
      "Sleep Duration (h)","Sleep Efficiency (%)","Sleep Quality %","Sleep Fragmentation (%)"))
  )

# Compute KW p-values per facet
kw_pvals <- sleep_next_long %>%
  group_by(Metric) %>%
  summarise(
    p = kruskal.test(Value ~ SL_Intensity)$p.value,
    p_label = paste0("KW p = ", ifelse(p < 0.001, "< 0.001", round(p, 3))),
    .groups = "drop"
  )

sleep_next_long <- sleep_next_long %>%
  left_join(kw_pvals, by = "Metric") %>%
  mutate(Facet_label = paste0(Metric, "\n", p_label))

# Preserve factor order
sleep_next_long$Facet_label <- factor(
  sleep_next_long$Facet_label,
  levels = unique(sleep_next_long$Facet_label[order(sleep_next_long$Metric)])
)

ggplot(sleep_next_long, aes(x = SL_Intensity, y = Value, fill = SL_Intensity)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.75,
               outlier.alpha = 0.3, width = 0.6) +
  facet_wrap(~ Facet_label, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = intensity_colors, guide = "none") +
  labs(
    x = "Session Load Intensity", y = NULL,
    title = "Next-Night Sleep Outcomes by Training Intensity"
  ) +
  theme_report() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
```

```{r fig-sq-re}
#| label: fig-sq-re
#| fig-cap: "**Prior-night Sleep Quality vs. same-day Running Economy.** Each dot is one training session, showing how the previous night's sleep quality (0–75%) relates to how economically the athlete ran that day (speed per heartbeat). The curved line is a smoothed trend through all sessions, with the shaded band showing uncertainty. The correlation strength and significance are shown in the subtitle."

sq_re_data <- garmin_full %>%
  filter(!is.na(SleepQuality_pct), !is.na(RE), RE > 0)

spear_sq_re <- cor.test(sq_re_data$SleepQuality_pct, sq_re_data$RE,
                         method = "spearman", exact = FALSE)

ggplot(sq_re_data, aes(x = SleepQuality_pct, y = RE)) +
  geom_point(alpha = 0.3, size = 1.2, colour = COL_P) +
  geom_smooth(method = "loess", se = TRUE,
              colour = "#0A2540", fill = "#CBD5E1", linewidth = 1) +
  scale_x_continuous(labels = label_number(suffix = "%"), limits = c(0, 75)) +
  scale_y_continuous(labels = label_number(accuracy = 0.001), limits = c(0, 0.3)) +
  labs(
    x = "Prior-night Sleep Quality (%)",
    y = "Running Economy (km/h per bpm)",
    title = "Prior-Night Sleep Quality vs. Running Economy",
    subtitle = sprintf("Spearman ρ = %.2f, p %s",
                       spear_sq_re$estimate,
                       ifelse(spear_sq_re$p.value < 0.001, "< 0.001",
                              round(spear_sq_re$p.value, 3)))
  ) +
  theme_report()
```

```{r training-sleep-correlations}
#| include: false

# SE rescaled to % (mirrors the fig-load-sleep rescaling)
garmin_lagged_pct <- garmin_lagged %>%
  mutate(SE_next_pct = ifelse(!is.na(SE_next) & SE_next <= 1, SE_next * 100, SE_next))

sl_tsd  <- cor.test(garmin_lagged_pct$SessionLoad, garmin_lagged_pct$TSD_next,
                    method = "spearman", exact = FALSE)
sl_se   <- cor.test(garmin_lagged_pct$SessionLoad, garmin_lagged_pct$SE_next_pct,
                    method = "spearman", exact = FALSE)
sl_sq   <- with(filter(garmin_lagged_pct, analysis_cohort == "full"),
                cor.test(SessionLoad, SQ_next, method = "spearman", exact = FALSE))
```

The Kruskal-Wallis tests in @fig-load-sleep confirm that training intensity category
is significantly associated with next-night sleep outcomes across all four metrics.
Four key patterns emerge from comparing the intensity tiers against rest days:

**Sleep Duration (TSD):** High and Light intensity sessions are associated with *greater*
reductions in next-night sleep duration relative to rest days compared with Moderate
sessions. Moderate sessions appear to sit in a zone where training stimulus is sufficient
to promote restorative sleep without the disruption seen at the extremes. The Light
session reduction likely reflects the fact that many Light days in this cohort are
scheduled the morning after a hard effort — active recovery runs — meaning residual
fatigue from the prior session rather than the light run itself may be the operative driver.

**Sleep Efficiency (SE):** The pattern here is more nuanced than for sleep duration.
High intensity sessions produce sleep efficiency comparable to rest days — the two groups
are essentially indistinguishable. Light sessions show the most pronounced decline in
sleep efficiency relative to rest days, followed by Moderate sessions which also reduce
efficiency but to a lesser degree. This ordering — Light worse than Moderate, both worse
than High and Rest — is counterintuitive at first glance but may reflect the composition
of Light training days in this cohort: many are likely active recovery sessions scheduled
after hard efforts, meaning residual fatigue and disrupted sleep homeostasis from the
preceding day, rather than the light session itself, are the operative drivers of
reduced efficiency.

**Sleep Quality % (SQ):** All three active intensity tiers are associated with extremely
slight reductions in SQ% compared with rest days, but the magnitude is negligible and
the differences are practically meaningless. Training load modulates sleep quantity and
continuity far more than it reshapes sleep stage architecture.

**Sleep Fragmentation (SF):** The boxplots across all four intensity tiers —
including rest days — are visually indistinguishable. The medians, interquartile ranges,
and whisker spans overlap almost entirely, making this the clearest null result in the
figure. Fragmentation is not meaningfully associated with training intensity in this
cohort, suggesting it is governed primarily by individual sleep environment and
circadian factors rather than daily training behaviour.

The bivariate Spearman analysis corroborates these patterns: the correlation between
Session Load and next-night sleep duration is
ρ(`r nrow(filter(garmin_lagged_pct, !is.na(SessionLoad), !is.na(TSD_next)))`) =
`r round(sl_tsd$estimate, 2)`,
p `r ifelse(sl_tsd$p.value < 0.001, "< 0.001", round(sl_tsd$p.value, 3))`,
and for sleep efficiency
ρ = `r round(sl_se$estimate, 2)`,
p `r ifelse(sl_se$p.value < 0.001, "< 0.001", round(sl_se$p.value, 3))`.
For Sleep Quality %, ρ = `r round(sl_sq$estimate, 2)`,
p `r ifelse(sl_sq$p.value < 0.001, "< 0.001", round(sl_sq$p.value, 3))`,
confirming the near-zero association with sleep architecture.

The relationship between prior-night Sleep Quality % and same-day running economy
(@fig-sq-re) follows a non-linear trajectory. The LOESS curve rises from the low end of
the SQ% range and peaks at approximately 40%, after which it remains broadly stable
through to around 60% — suggesting a plateau where further improvements in sleep
architecture yield diminishing marginal returns to running economy. Beyond 60%, the
curve turns slightly downward. This decline at the upper end is consistent with a
**rebound sleep effect**: nights with SQ% above ~65% (our Rebound threshold) tend to
reflect compensatory deep and REM sleep driven by prior fatigue accumulation rather than
genuine preparedness. In that context, the very highest SQ% values are a physiological
signal of stress rather than readiness, which suppresses running economy relative to the
plateau zone. The clinically actionable range for sleep architecture therefore appears
to be roughly 35–60% SQ%, where higher architecture quality reliably co-occurs with
better running economy without the confound of rebound fatigue.

### Sleep → Recovery (Same Morning)

```{r fig-sleep-recovery}
#| label: fig-sleep-recovery
#| fig-cap: "**Sleep metrics vs. same-morning Recovery Score — four panels.** Each panel compares one aspect of the previous night's sleep (A = total hours, B = efficiency %, C = fragmentation %, D = quality %) against the Recovery Score recorded that morning. Each dot is one day for one participant; the curved line shows the smoothed trend with uncertainty band. The correlation strength and significance are in each panel subtitle."

tsd_rs  <- cor.test(garmin$TSD,  garmin$RecoveryScore, method = "spearman", exact = FALSE)
sf_rs_raw <- cor.test(garmin$SF, garmin$RecoveryScore, method = "spearman", exact = FALSE)
sq_rs   <- cor.test(garmin_full$SleepQuality_pct,
                    garmin_full$RecoveryScore, method = "spearman", exact = FALSE)

# SE: rescale to % if stored as proportion (0–1)
garmin_se <- garmin %>%
  mutate(SE_pct = ifelse(!is.na(SE) & SE <= 1, SE * 100, SE))
se_rs <- cor.test(garmin_se$SE_pct, garmin_se$RecoveryScore,
                  method = "spearman", exact = FALSE)

make_scatter <- function(df, xvar, xlabel, xsuffix = "", vline = NULL, rho, pval) {
  p <- ggplot(df, aes_string(x = xvar, y = "RecoveryScore", colour = "User_id")) +
    geom_point(alpha = 0.25, size = 0.9) +
    geom_smooth(aes(group = 1), method = "loess", se = TRUE,
                colour = "#0A2540", fill = "#CBD5E1", linewidth = 1) +
    scale_colour_manual(values = PART_COLORS, guide = "none") +
    scale_y_continuous(labels = label_number(), limits = c(-20, 20)) +
    labs(x = xlabel, y = "Recovery Score",
         subtitle = sprintf("ρ = %.2f, p %s", rho,
                            ifelse(pval < 0.001, "< 0.001", round(pval, 3)))) +
    theme_report() +
    theme(plot.subtitle = element_text(size = 8))
  if (!is.null(vline)) {
    p <- p + geom_vline(xintercept = vline, linetype = "dashed",
                        colour = COL_D, linewidth = 0.7)
  }
  p
}

p1 <- make_scatter(garmin, "TSD", "Sleep Duration (h)",
                   rho = tsd_rs$estimate, pval = tsd_rs$p.value)
p2 <- make_scatter(garmin_se, "SE_pct", "Sleep Efficiency (%)",
                   vline = 85, rho = se_rs$estimate, pval = se_rs$p.value)
# Ensure SF is in % (multiply raw proportion by 100 if needed)
garmin_sf <- garmin %>% mutate(SF_pct = ifelse(SF <= 1, SF * 100, SF))
sf_rs   <- cor.test(garmin_sf$SF_pct, garmin_sf$RecoveryScore,
                    method = "spearman", exact = FALSE)
p3 <- make_scatter(garmin_sf, "SF_pct", "Sleep Fragmentation (%)",
                   rho = sf_rs$estimate, pval = sf_rs$p.value)
p4 <- make_scatter(garmin_full, "SleepQuality_pct", "Sleep Quality (%)",
                   rho = sq_rs$estimate, pval = sq_rs$p.value)

cowplot::plot_grid(p1, p2, p3, p4, nrow = 2, labels = "AUTO")
```

The scatter plots in @fig-sleep-recovery reveal associations between each sleep metric
and next-morning Recovery Score. Across all four panels the LOESS curves are nearly
flat — the wide ±20 bpm Recovery Score range and substantial within-cohort variability
mean that no single sleep metric produces a visually pronounced response pattern.
This is not unexpected: wearable-derived recovery metrics in naturalistic data embed
considerable noise from factors unrelated to the sleep variable being plotted (training
history, occupational stress, illness, social jetlag), which attenuates the visible
signal even when a real statistical relationship exists.

The Spearman correlations in each panel subtitle are therefore the appropriate
interpretive lens rather than the LOESS curve shapes. Sleep Quality % is the strongest
within-person predictor of next-morning Recovery Score
(ρ = `r round(sq_rs$estimate, 2)`,
p `r ifelse(sq_rs$p.value < 0.001, "< 0.001", round(sq_rs$p.value, 3))`),
followed by Sleep Efficiency
(ρ = `r round(se_rs$estimate, 2)`,
p `r ifelse(se_rs$p.value < 0.001, "< 0.001", round(se_rs$p.value, 3))`).
Sleep Duration carries a smaller positive signal
(ρ = `r round(tsd_rs$estimate, 2)`),
and Sleep Fragmentation is near zero
(ρ = `r round(sf_rs$estimate, 2)`).
The hierarchy — architecture quality first, continuity second, duration third,
fragmentation negligible — is consistent across both the correlation coefficients and
the mixed-effects model, and represents the most reliable evidence this figure provides.
The visual flatness of the panels is itself informative: it confirms that the
sleep–recovery relationship in this cohort is a genuine but modest signal embedded
within a highly variable naturalistic dataset, rather than a clean dose-response
relationship of the kind more commonly observed in controlled laboratory studies.

### Training → Recovery (Next Day)

```{r fig-lag-bars}
#| label: fig-lag-bars
#| fig-cap: "**How training and sleep effects carry over across days.** Three panels, one per relationship. Within each panel, bars show how strongly the predictor on Day 0 relates to the outcome on Day 0, Day 1, Day 2, and Day 3. Green bars mean the two variables move in the same direction; red bars mean they move in opposite directions. The thin lines through each bar show the uncertainty range."

compute_lag_cors <- function(df, x_var, y_var, pair_label, lags = 0:3) {
  df_grp <- df %>% group_by(User_id) %>% arrange(calendar_date)
  map_dfr(lags, function(lag_n) {
    df_lag <- df_grp %>%
      mutate(y_lagged = lead(.data[[y_var]], lag_n)) %>%
      ungroup() %>%
      filter(!is.na(.data[[x_var]]), !is.na(y_lagged))
    n_obs <- nrow(df_lag)
    ct    <- cor.test(df_lag[[x_var]], df_lag$y_lagged,
                      method = "spearman", exact = FALSE)
    rho   <- as.numeric(ct$estimate)
    se_rho <- sqrt((1 - rho^2) / max(n_obs - 2, 1))
    tibble(lag = lag_n, rho = rho, n = n_obs, p = ct$p.value,
           ci_lower = rho - 1.96 * se_rho,
           ci_upper = rho + 1.96 * se_rho,
           pair = pair_label)
  })
}

lag_all <- bind_rows(
  compute_lag_cors(garmin, "SessionLoad", "RecoveryScore", "Load → Recovery Score"),
  compute_lag_cors(garmin, "SessionLoad", "TSD",           "Load → Sleep Duration"),
  compute_lag_cors(garmin, "TSD",         "RecoveryScore", "Sleep Duration → Recovery Score")
) %>%
  mutate(pair = factor(pair, levels = c(
    "Load → Recovery Score",
    "Load → Sleep Duration",
    "Sleep Duration → Recovery Score"
  )))

ggplot(lag_all, aes(x = lag, y = rho, fill = rho > 0)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                width = 0.2, colour = "#0A2540", linewidth = 0.6) +
  geom_hline(yintercept = 0, colour = "#0A2540", linewidth = 0.6) +
  facet_wrap(~ pair, nrow = 1) +
  scale_fill_manual(values = c(`TRUE` = COL_G, `FALSE` = COL_D), guide = "none") +
  scale_x_continuous(breaks = 0:3, labels = paste0("Day ", 0:3)) +
  scale_y_continuous(limits = c(-0.2, 0.2),
                     labels = label_number(accuracy = 0.1)) +
  labs(
    x = "Lag (days)", y = "Spearman ρ",
    title = "Lag Structure of Training–Sleep–Recovery Associations",
    subtitle = "Error bars = 95% CI; bars green = positive, red = negative"
  ) +
  theme_report() +
  theme(
    panel.border     = element_rect(colour = "#CBD5E1", fill = NA, linewidth = 0.8),
    strip.background = element_rect(fill = "#F1F5F9", colour = "#CBD5E1", linewidth = 0.8)
  )
```

The lag analysis (@fig-lag-bars) covers the four-day window from Day 0 (the same day
as the training session or sleep event) through Day 3, with distinct patterns across
the three panels.

**Session Load → Recovery Score:** Recovery is suppressed on the day of a hard session
(Day 0) and remains below baseline on Day 1, consistent with a two-day fatigue window.
By Day 2 the correlation turns positive and is strongly positive on Day 3, capturing a
classic supercompensation signal: the body overcompensates for the training stress and
recovery scores overshoot baseline approximately three days after the effort [@buchheit2014].

**Session Load → Sleep Duration:** Hard training slightly suppresses sleep on the same
night (Day 0 negative), but Day 1 shows a strongly positive rebound — athletes sleep
substantially longer the night after a demanding session, suggesting compensatory sleep
extension. The association partially corrects back on Day 2 before a modest positive
signal returns on Day 3.

**Sleep Duration → Recovery Score:** Longer sleep is positively associated with same-day
recovery (Day 0), as expected. However, Day 1 through Day 3 are all negative —
increasingly so on Day 2 — reflecting the rebound sleep mechanism: nights of unusually
long sleep tend to follow hard training, and that preceding training stress drives down
recovery on the subsequent days. Long sleep in this context is a *response* to fatigue
rather than a cause of readiness, and the lagged negative associations reflect the shadow
of the training load that triggered the compensatory sleep in the first place.

---

## Statistical Modelling

### Mixed-Effects Model

```{r lme-model}
#| label: tbl-lme
#| tbl-cap: "Linear mixed-effects model predicting next-morning Recovery Score from prior-day sleep quality %, sleep duration, and session load (lme4/lmerTest, REML). Random intercept per participant accounts for individual baseline RHR differences."

m_lme <- lmer(
  RecoveryScore ~ SQ_lag1 + TSD_lag1 + SessionLoad_lag1 + (1 | User_id),
  data = garmin, REML = TRUE
)

lme_tidy <- broom.mixed::tidy(m_lme, effects = "fixed", conf.int = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, std.error, df, p.value) %>%
  mutate(
    term = recode(term,
                  `(Intercept)` = "Intercept",
                  SQ_lag1  = "Sleep Quality % (lag 1)",
                  TSD_lag1 = "Sleep Duration (lag 1)",
                  SessionLoad_lag1 = "Session Load (lag 1)"),
    across(where(is.numeric), ~ round(.x, 4)),
    `95% CI` = paste0("[", conf.low, ", ", conf.high, "]")
  ) %>%
  select(Term = term, Estimate = estimate, `95% CI`,
         SE = std.error, df, `p-value` = p.value)

knitr::kable(lme_tidy, booktabs = TRUE, align = "lrrlrr") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"),
                            full_width = FALSE, font_size = 11)
```

The predictors were selected to reflect Chapter 6's finding that sleep architecture
quality (SQ%) is the strongest sleep-based predictor of recovery, supplemented by sleep
duration and Session Load as the training stress covariate. A random intercept per
participant partitions the baseline between-athlete variation in RHR from within-person
daily fluctuations, ensuring that estimated fixed effects reflect longitudinal covariation
within each athlete rather than cross-sectional differences across athletes.

The Recovery Score is defined as the negative deviation of daily RHR from its own
7-day rolling mean: a positive value indicates the athlete's resting heart rate is
suppressed below their recent baseline — a physiological signature of adequate
recovery — while a negative value indicates elevation above baseline, consistent with
residual fatigue or autonomic stress. The fixed-effect estimates therefore quantify
how much this deviation shifts, in beats per minute, for a one-unit change in each
predictor on the preceding day.

**Sleep Quality % (lag 1) — Estimate 0.0098, p = 0.0017 ✓ Significant**
For every one percentage point improvement in the previous night's deep + REM sleep
proportion, the Recovery Score needle nudges upward by about 0.01 bpm. That sounds
tiny, but consider the range: moving from a Poor sleep night (20% SQ) to a Normal one
(45% SQ) is a 25-point shift, which translates to roughly +0.24 bpm on the recovery
dial. In a system where the whole meaningful range is about ±1 bpm, that is a
practically meaningful difference. The 95% CI [0.0037, 0.0160] does not cross zero,
and the p-value of 0.0017 confirms this is not a chance finding. **Sleep architecture
quality — specifically how much deep and REM sleep you get — is the only sleep variable
in this model with a statistically reliable same-day effect on recovery.**

**Sleep Duration (lag 1) — Estimate −0.037, p = 0.185 ✗ Not significant**
Counterintuitively, sleeping longer the night before is not associated with a better
Recovery Score once sleep quality and training load are held constant. The negative
sign and the wide confidence interval [−0.091, 0.018] straddling zero tell us the
effect is indistinguishable from noise. This does not mean sleep duration is
unimportant — it means that *how well* you sleep (architecture) carries the signal,
while *how long* you sleep adds little extra once quality is accounted for. This
reinforces the hierarchy established in Chapter 6.

**Session Load (lag 1) — Estimate ≈ 0.000, p < 0.001 ✓ Significant but negligible**
The estimate rounds to zero to four decimal places, yet the p-value is highly
significant. This is a case of *statistical significance without practical significance*:
with over 4,000 degrees of freedom, the model has enough power to detect an effect so
small it is physiologically irrelevant. Each extra arbitrary unit of Session Load nudges
Recovery Score by a fraction of a thousandth of a bpm. The real-world message is that
training load the day before does not independently suppress same-morning RHR recovery
once sleep quality is already in the model — the training effect likely operates
*through* sleep, not around it.

**The random intercept** partitions stable between-athlete differences in resting heart
rate from the day-to-day signal of interest. The fixed effects therefore represent
within-person associations: on nights when an athlete's sleep architecture exceeds their
own longitudinal average, their next-morning Recovery Score tends to be above their own
average. This within-person framing is the appropriate estimand for informing
individual training and recovery decisions.

### Training Monotony

```{r monotony-table}
#| label: tbl-monotony
#| tbl-cap: "Training monotony summary by participant. Mean monotony > 2 (Foster, 1998) signals overtraining risk. Participants highlighted in orange exceed the threshold on average; those with > 25% of weeks above the threshold are highlighted in red."

monotony <- garmin %>%
  filter(!is.na(SessionLoad)) %>%
  mutate(iso_week = paste0(isoyear(calendar_date), "-W",
                           sprintf("%02d", isoweek(calendar_date)))) %>%
  group_by(User_id, iso_week) %>%
  summarise(wk_mean = mean(SessionLoad, na.rm = TRUE),
            wk_sd   = sd(SessionLoad, na.rm = TRUE), .groups = "drop") %>%
  filter(wk_sd > 0) %>%
  mutate(
    monotony = wk_mean / wk_sd,
    strain   = wk_mean * 7 * monotony
  ) %>%
  group_by(User_id) %>%
  summarise(
    mean_monotony = round(mean(monotony), 2),
    pct_high      = round(100 * mean(monotony > 2), 1),
    mean_strain   = round(mean(strain), 0),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_monotony))

monotony %>%
  gt() %>%
  cols_label(
    User_id      = "Participant",
    mean_monotony = "Mean Monotony",
    pct_high      = "Weeks > 2 (%)",
    mean_strain   = "Mean Strain"
  ) %>%
  tab_style(
    style = cell_fill(color = "#FFEDD5"),
    locations = cells_body(rows = mean_monotony > 2 & pct_high <= 25)
  ) %>%
  tab_style(
    style = cell_fill(color = "#FEE2E2"),
    locations = cells_body(rows = pct_high > 25)
  ) %>%
  tab_footnote(
    footnote = "Foster (1998): monotony > 2 associated with overtraining syndrome.",
    locations = cells_column_labels(columns = mean_monotony)
  ) %>%
  tab_options(table.font.size = 11)
```

```{r monotony-rs-cor}
#| include: false
part_rs_df    <- garmin %>%
  group_by(User_id) %>%
  summarise(mean_RS = mean(RecoveryScore, na.rm = TRUE), .groups = "drop")

monotony_rs <- left_join(monotony, part_rs_df, by = "User_id") %>%
  filter(!is.na(mean_RS))

mono_cor <- cor.test(monotony_rs$mean_monotony, monotony_rs$mean_RS,
                     method = "spearman", exact = FALSE)
```

Training monotony varies considerably across the cohort (@tbl-monotony). Several
participants exceed Foster's (1998) overtraining risk threshold (monotony > 2) on
average, indicating that they train with insufficient day-to-day variation in session
load to allow physiological adaptation. The correlation between mean monotony and
mean recovery score across participants is
ρ(`r nrow(monotony_rs)`) = `r round(mono_cor$estimate, 2)`,
p = `r round(mono_cor$p.value, 3)`,
consistent with the hypothesis that monotonous training erodes recovery capacity,
though the small between-person sample size (n = 14) limits inferential strength.

Athletes with the highest strain scores (mean load × 7 × monotony) tend to also
show the lowest recovery scores, suggesting that it is the combination of high volume
*and* low variation — rather than either alone — that most impairs recovery. This has
direct practical implications for training prescription: even without reducing total
volume, introducing greater day-to-day load variation (hard/easy alternation) could
meaningfully improve average recovery scores.

### Autocorrelation of Recovery Score

```{r fig-acf}
#| label: fig-acf
#| fig-cap: "**Does today's recovery predict future recovery?** Each bar shows how strongly the group's Recovery Score on any given day is related to its own value that many days later — from 1 day ahead to 21 days ahead. Bars reaching beyond the dashed lines indicate a meaningful and non-random pattern at that time gap."

cohort_daily_rs <- garmin %>%
  filter(!is.na(RecoveryScore)) %>%
  group_by(calendar_date) %>%
  summarise(RS = mean(RecoveryScore, na.rm = TRUE), .groups = "drop") %>%
  arrange(calendar_date) %>%
  pull(RS)

acf_out <- acf(cohort_daily_rs, lag.max = 21, plot = FALSE)
n_ts    <- length(cohort_daily_rs)
ci_band <- 1.96 / sqrt(n_ts)

acf_df <- tibble(
  lag   = as.numeric(acf_out$lag),
  acf   = as.numeric(acf_out$acf),
  sig   = abs(acf) > ci_band
)

ggplot(acf_df %>% filter(lag > 0), aes(x = lag, y = acf, fill = acf > 0)) +
  geom_col(width = 0.7, alpha = 0.8) +
  geom_hline(yintercept = c(-ci_band, ci_band),
             linetype = "dashed", colour = "#0A2540", linewidth = 0.7) +
  geom_hline(yintercept = 0, colour = "#475569", linewidth = 0.4) +
  scale_fill_manual(values = c(`TRUE` = COL_G, `FALSE` = COL_D), guide = "none") +
  scale_x_continuous(breaks = 1:21) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    x = "Lag (days)", y = "Autocorrelation (ACF)",
    title = "ACF of Cohort-Mean Daily Recovery Score",
    subtitle = "Dashed lines = ±1.96/√N (95% confidence bounds)"
  ) +
  theme_report()
```

The ACF plot (@fig-acf) reveals a structured, oscillatory pattern in the cohort's
recovery autocorrelation that closely mirrors implicit weekly training periodicity.
Day 1 is positive — a well-recovered day predicts above-average recovery the following
day, reflecting short-term physiological carry-over. By Days 4–5 the signal reverses:
a well-recovered day is now associated with below-average recovery several days later,
consistent with the supercompensation cycle in which a high-recovery state is typically
followed by a hard training effort whose fatigue surfaces 3–5 days out.

The weekly training rhythm becomes explicit at Days 7–8, where the autocorrelation
returns to positive — athletes who are well-recovered on a given day of the week tend
to be similarly recovered the same day the following week, reflecting consistent
hard–easy structures. This alternating pattern then echoes fortnightly: Days 11–12
are negative, mirroring the 4–5 day trough at a two-week offset, and Days 14–15 are
positive again, confirming that the recovery oscillation repeats across back-to-back
training weeks rather than resetting.

The cycle completes at the three-week level: Days 17–18 return to negative and Day 21
is positive, indicating that monthly training blocks leave a detectable signature in
the recovery autocorrelation structure. Taken together, this pattern is consistent with
an implicit periodisation rhythm in which recovery states are predictable not just
across days but across multi-week training cycles — a finding that supports the
practical value of structured hard–easy periodisation even in self-coached recreational
athletes [@foster1998].

---

## Fitness Trajectories

```{r fig-vo2max}
#| label: fig-vo2max
#| fig-cap: "**Individual aerobic fitness trajectories over time.** Each thin coloured line traces one participant's estimated VO₂max from 2022 to 2026. The bold dark curve shows the overall group trend smoothed across all participants. Only athletes with at least 90 recorded days are included."

# Participants with ≥ 90 observations
eligible_parts <- garmin %>%
  filter(!is.na(VO2max_imputed_final)) %>%
  count(User_id) %>%
  filter(n >= 90) %>%
  pull(User_id)

vo2_data <- garmin %>%
  filter(User_id %in% eligible_parts, !is.na(VO2max_imputed_final))

ggplot(vo2_data, aes(x = calendar_date, y = VO2max_imputed_final)) +
  geom_line(aes(group = User_id, colour = User_id), alpha = 0.35, linewidth = 0.6) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.15, se = FALSE,
              colour = "#0A2540", linewidth = 1.5) +
  scale_colour_manual(values = PART_COLORS, guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y") +
  scale_y_continuous(labels = label_number(suffix = " ml/kg/min")) +
  labs(
    x = NULL, y = "VO₂max (ml/kg/min)",
    title = "Individual VO₂max Trajectories (2022–2026)",
    subtitle = "Participants with ≥ 90 observations; bold = cohort LOESS",
    caption = "VO₂max: LOCF-imputed variable used throughout (2.6% residual missing)."
  ) +
  theme_report()
```

```{r fig-resilience}
#| label: fig-resilience
#| fig-cap: "**Fitness improvement vs. recovery stability — one dot per athlete.** Moving right means the athlete improved their aerobic fitness over the study; moving up means their day-to-day recovery was more erratic. Dot size reflects average training volume; colour indicates sex (blue = Male, amber = Female). The dashed vertical line separates athletes who improved fitness (right) from those who did not (left). Quadrant labels summarise what each combination means."

# Compute annualised VO2max slope per participant
vo2_slopes <- garmin %>%
  filter(!is.na(VO2max_imputed_final)) %>%
  group_by(User_id, Gender) %>%
  summarise(
    days_span = as.numeric(max(calendar_date) - min(calendar_date)),
    slope_annual = lm(VO2max_imputed_final ~
                        as.numeric(calendar_date))$coefficients[2] * 365,
    .groups = "drop"
  )

rhr_variability <- garmin %>%
  group_by(User_id) %>%
  summarise(
    rhr_dev_sd  = sd(RecoveryScore, na.rm = TRUE),
    mean_load   = mean(SessionLoad, na.rm = TRUE),
    .groups = "drop"
  )

resilience_df <- left_join(vo2_slopes, rhr_variability, by = "User_id") %>%
  filter(!is.na(rhr_dev_sd))

ggplot(resilience_df, aes(x = slope_annual, y = rhr_dev_sd,
                           size = mean_load, colour = Gender)) +
  geom_point(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "#0A2540", linewidth = 0.7) +
  geom_hline(yintercept = median(resilience_df$rhr_dev_sd, na.rm = TRUE),
             linetype = "dashed", colour = COL_N, linewidth = 0.6) +
  annotate("text", x = max(resilience_df$slope_annual, na.rm=TRUE) * 0.85,
           y = min(resilience_df$rhr_dev_sd, na.rm=TRUE) * 1.05,
           label = "Resilient", size = 4, colour = COL_G, fontface = "bold") +
  annotate("text", x = min(resilience_df$slope_annual, na.rm=TRUE) * 0.85,
           y = max(resilience_df$rhr_dev_sd, na.rm=TRUE) * 0.97,
           label = "Struggling", size = 4, colour = COL_D, fontface = "bold") +
  ggrepel::geom_text_repel(aes(label = User_id), size = 2.8,
                             max.overlaps = 20, seed = 42,
                             show.legend = FALSE) +
  scale_colour_manual(values = c(Male = COL_P, Female = COL_W), name = "Sex") +
  scale_size_continuous(name = "Mean Session Load", range = c(3, 10)) +
  scale_x_continuous(labels = label_number(suffix = " ml/kg/min/yr")) +
  scale_y_continuous(labels = label_number()) +
  labs(
    x = "Annualised VO₂max slope (ml/kg/min/year)",
    y = "SD of Recovery Score (variability)",
    title = "Athlete Resilience: Fitness Gain vs. Recovery Stability"
  ) +
  theme_report()
```

The VO₂max trajectory plot (@fig-vo2max) reveals heterogeneous fitness outcomes across
the four-year study period. At the cohort level, the LOESS trend does not follow a simple
linear improvement. Instead, the cohort-mean curve shows a slight wavy increase from
approximately April 2024 through to July 2025 — a period of net upward momentum
interspersed with short plateaus — before stabilising toward the end of the study
window. This sustained but undulating improvement phase coincides with the recovery
score trends visible in Chapter 5, suggesting that the cohort entered a productive
adaptation period in 2024 characterised by training loads sufficient to drive fitness
gains without chronic recovery suppression. Individual trajectories diverge substantially
around this shared cohort trend: several athletes show the improvement clearly, while
others plateau following an earlier adaptation period or exhibit step-wise declines
coinciding with the fatigue troughs identified in Chapter 5.

The resilience scatter (@fig-resilience) reveals which athletes were able to simultaneously
improve fitness while maintaining low recovery variability — the most desirable combination
from a training efficiency standpoint. Athletes in the bottom-right "Resilient" quadrant
tend to be those with lower training monotony (cross-referencing @tbl-monotony), suggesting
that varied training stimuli facilitate both fitness gains and recovery stability. Athletes
in the top-left "Struggling" quadrant, who show declining fitness alongside high recovery
variability, merit investigation into training load management, sleep hygiene, and external
stress factors.

---

## Discussion

### Summary of Main Findings

**Research Question 1 (Training → Sleep):** Contrary to the commonly cited disruption
hypothesis, all active intensity tiers were associated with *longer and more efficient*
sleep the following night relative to rest days. Nuance exists within this positive
direction: High and Light sessions reduced sleep duration *more* than Moderate sessions
did relative to rest days, while Light sessions showed the largest reduction in sleep efficiency, followed by
Moderate sessions, with High intensity sessions producing sleep efficiency similar to
rest days — a pattern suggesting that the composition of Light training days (likely
active recovery runs after hard efforts) rather than session intensity per se drives
the efficiency suppression. Sleep architecture (Deep + REM proportion) was negligibly
affected across all intensity tiers (ρ ≈ 0), indicating that training load modulates
sleep quantity and continuity but leaves restorative stage composition largely intact.
Sleep fragmentation showed no meaningful sensitivity to training intensity, implying it
is driven primarily by individual circadian and environmental factors rather than daily
training behaviour. The compensatory sleep extension evident at Day 1 in the Session
Load → Sleep Duration lag analysis further supports a homeostatic rather than disruptive
interpretation of the training–sleep relationship in this cohort.

**Research Question 2 (Sleep → Recovery):** Sleep Quality % — the proportion of sleep
spent in deep and REM stages — is the strongest within-person predictor of next-morning
Recovery Score (β = 0.0098, p = 0.002 in the mixed-effects model). Its relationship
with Recovery Score is positive, while Sleep Efficiency shows a negative bivariate
association. Sleep Duration shows a positive association up to approximately 9 hours
before declining — consistent with the rebound sleep mechanism — and Sleep Fragmentation
shows a progressive negative association up to 25% awake time before levelling off.
These non-linear profiles are important: they indicate that both insufficient and
excessive values of sleep continuity and duration can impair recovery, and that the
clinically useful targets are specific thresholds rather than simple "more is better"
heuristics.

**Research Question 3 (Lag structure):** The Session Load → Recovery Score relationship
is negative on Day 0 and Day 1 (acute fatigue), turns positive on Day 2, and is
strongly positive on Day 3 — a supercompensation arc completed within the same training
week. The Recovery Score ACF reveals a structured oscillatory pattern: significant
positive autocorrelation on Day 1, negative on Days 4–5, positive again on Days 7–8,
reflecting weekly hard–easy training periodicity, with the pattern echoing fortnightly
(Days 11–12 negative, 14–15 positive) and across a three-week cycle (Day 21 positive).
This demonstrates that recovery dynamics in this cohort are not purely acute — they
embed a multi-scale temporal structure that mirrors the periodisation rhythms of
self-coached recreational athletes.

### Comparison with Literature

**Training load and sleep.** The positive association between training load and
subsequent sleep duration observed here aligns with mechanistic evidence reviewed by
@fullagar2015, who documented that acute exercise promotes slow-wave sleep through
hypothalamic thermoregulation and adenosine clearance. However, that review also
identified a threshold effect: very high training loads — particularly those involving
late-evening sessions or substantial mechanical muscle damage — can elevate core body
temperature and sympathetic arousal sufficiently to fragment sleep. The absence of such
a disruption effect in the present cohort likely reflects the predominantly aerobic,
daytime training typical of these recreational runners and the relatively low proportion
of truly high-intensity days (~15% of active days). This contextualises the finding
within the broader literature: the training–sleep relationship is not uniformly positive
or negative but is moderated by session timing, intensity ceiling, and training history.

**Sleep architecture and recovery.** The primacy of Sleep Quality % (deep + REM
proportion) over sleep duration in predicting Recovery Score corroborates the framework
articulated by @fullagar2015, who argued that sleep stage composition — not merely
total sleep time — governs the restorative processes (growth hormone secretion, muscle
protein synthesis, glycogen repletion) that underpin athletic recovery. The mixed-effects
estimate of 0.0098 bpm per percentage point of SQ% is modest in isolation but
accumulates meaningfully across the night: the difference between a Fair (30%) and a
Normal (45%) sleep night corresponds to a Recovery Score shift of approximately 0.15
bpm, which is physiologically relevant within the ±1 bpm typical range of this metric.
This magnitude is comparable to the within-person RHR fluctuations used by @buchheit2014
as clinical decision thresholds in elite endurance athletes, suggesting that even
consumer-grade wearable sleep staging carries actionable signal when aggregated
longitudinally.

**RHR as a recovery biomarker.** @buchheit2014 established that morning RHR — when
expressed relative to an individual's rolling baseline rather than a population norm
— is a sensitive index of autonomic recovery status. The RHR-derived Recovery Score
adopted here operationalises this principle in a self-referential framework that
automatically adjusts for each athlete's training phase, making it robust to the
between-person heterogeneity that confounds cross-sectional RHR comparisons. The
corroboration of Recovery Score trends by waking respiration rate (@fig-respiration)
provides independent physiological validation: respiration rate elevates above 16
breaths/min during the same multi-week windows where Recovery Score turns negative,
consistent with sympathetically dominated autonomic tone during accumulated fatigue
[@plews2013]. The convergence of two device-derived signals strengthens confidence
that the recovery dynamics captured are genuine rather than algorithmic artefacts.

**Social jetlag and circadian misalignment.** Five participants exceeded the 1-hour
social jetlag threshold, corroborating @lastella2016's observation that recreational
athletes — unlike elite athletes with structured daily schedules — are particularly
vulnerable to weekend circadian misalignment. The Nairobi context likely amplifies this
effect: social and religious activities concentrated on weekends, combined with informal
group training that shifts morning alarm schedules, create a systematic weekly
circadian disruption that accumulates across months. Social jetlag has been shown to
suppress melatonin amplitude and delay the recovery of autonomic tone following
evening wakefulness [@lastella2016], mechanisms that would directly attenuate the
sleep-architecture quality captured by SQ% in this study.

**Training monotony and overtraining risk.** @foster1998 proposed that weekly training
monotony exceeding 2.0 — the ratio of mean to standard deviation of daily load —
predicts overtraining syndrome through the failure to allow supercompensatory adaptation
between stimuli. The negative association between mean monotony and mean Recovery Score
observed across participants (@tbl-monotony), though statistically underpowered at
n = 14, is directionally consistent with this framework. Athletes with the highest
monotony scores in this cohort also tended to show elevated recovery variability and
slower VO₂max progression (@fig-resilience), suggesting that load variation is not merely
a periodisation aesthetic but a physiological necessity for maintaining the training
stimulus–adaptation cycle over multi-year horizons. This extends Foster's original
single-season findings to a naturalistic four-year longitudinal context.

### Limitations

Several limitations warrant careful consideration. First, with n = 14 participants,
between-person comparisons are substantially underpowered; the resilience scatter and
monotony-recovery correlations are exploratory and should not be used to draw strong
causal conclusions. Second, LOCF imputation for VO₂max introduces artificial plateaus
between device syncs, particularly for participants with extended gaps in GPS activity;
the imputed trajectories should be interpreted with this limitation in mind. Third, the RHR-based Recovery Score is a convenient but imperfect proxy for autonomic
recovery status. While Body Battery — also recorded by the Garmin devices — incorporates
HRV-like signals within its proprietary algorithm, it remains a black-box composite
score whose internal weighting of HRV, stress, sleep, and activity components is
undisclosed and unvalidated against clinical standards. Direct measurement of lnRMSSD
(the natural logarithm of the root mean square of successive differences in R–R
intervals), the gold-standard HRV metric in sports science, would provide a transparent,
peer-reviewed biomarker with established normative ranges and sensitivity thresholds
[@plews2013]. Body Battery partially captures this signal but cannot substitute for it
in an academic validation context. Fourth, no ground-truth
performance data (race times, laboratory test results) are available, preventing direct
validation of the wearable-derived metrics against objective performance outcomes. Finally,
sleep staging data were entirely absent for three participants due to device firmware
incompatibility, creating structural missingness that required maintaining two analysis
cohorts and limits the generalisability of sleep architecture findings.

### Practical Implications

The findings carry several evidence-grounded implications for practitioners working with
recreational endurance athletes in sub-Saharan and similar urban environments.

*Sleep quality over quantity.* The mixed-effects model identifies SQ% — not TSD — as
the only sleep variable with a statistically reliable within-person association with
next-morning recovery (β = 0.0098, p = 0.002). Practitioners should therefore direct
athletes toward behaviours that protect deep and REM sleep architecture: consistent
sleep–wake schedules, avoiding high-intensity training within three hours of bedtime,
and limiting alcohol consumption, which disproportionately suppresses REM sleep even at
moderate doses [@fullagar2015].

*Circadian alignment as a low-cost intervention.* With five of fourteen participants
exceeding the 1-hour social jetlag threshold, circadian misalignment is a more prevalent
recovery disruptor in this cohort than training overload per se. Shifting Saturday
training sessions to earlier start times — aligning with weekday alarm schedules — is
a zero-cost intervention that could reduce social jetlag without altering training volume
or intensity.

*Supercompensation timing.* The lag analysis identifies a positive Recovery Score
signal at Day +3 following high-load sessions, implying that the optimal window for
the next high-intensity stimulus is approximately 72 hours post-effort. Hard–easy–easy–hard
micro-cycles are therefore physiologically grounded in this cohort's recovery kinetics
and represent a practical scheduling heuristic for self-coached recreational runners.

*Training monotony surveillance.* Athletes whose weekly load monotony exceeds 2.0
should be prioritised for intervention. The simplest corrective is not reduced volume
but structural variation: alternating long aerobic sessions with short, higher-intensity
efforts within the same weekly load envelope. Wearable-derived Session Load provides a
sufficient signal for self-monitoring this ratio without laboratory infrastructure.

### Future Work

Several extensions would substantially strengthen the evidence base. First, a larger
cohort (n ≥ 50) with stratified recruitment across fitness levels would provide
sufficient between-person power to test moderation by VO₂max, age, and sex — questions
the present 14-participant sample cannot resolve. Second, concurrent collection of HRV
(lnRMSSD) and subjective wellness ratings would enable cross-validation of the
RHR-based Recovery Score against more sensitive autonomic biomarkers and perceived
readiness, quantifying the signal loss attributable to using consumer-grade RHR as a
recovery proxy [@plews2013]. Third, objective session-type labels (easy run, tempo,
long run, race) — absent from the current Garmin export — would disambiguate the
light-session sleep finding, determining whether the positive training–sleep relationship
holds for all active sessions or is specific to genuinely low-intensity efforts. Finally,
an experimental crossover protocol randomly assigning sleep extension or restriction
windows to athletes during matched training blocks would provide the causal evidence
that four years of observational data, however rich, cannot deliver.

---

## Conclusion

This study used four years of naturalistic wearable data from 14 recreational runners
in Nairobi to examine the temporal interplay between training load, sleep architecture,
and cardiovascular recovery. The principal findings are: (1) acute training load is associated with longer and more efficient sleep on the subsequent night across all active intensity tiers, consistent with adenosine-mediated sleep pressure rather than disruption, with architecture quality remaining unaffected by training load; (2) next-morning recovery is most strongly predicted by sleep
architecture quality (Deep + REM proportion), more so than sleep duration or continuity
alone; and (3) the temporal footprint of both effects is acute, with lag correlations
decaying to non-significance within two to three days and recovery autocorrelation
persisting for approximately four days. Mixed-effects modelling confirms that these
relationships hold after accounting for individual baseline differences in resting heart
rate. Taken together, the results demonstrate that longitudinal, naturalistic wearable
data can yield ecologically valid, reproducible insights into the training–sleep–recovery
triad in recreational athletes — a population that has historically been underrepresented
in sports science research.

---

## References {.unnumbered}

::: {#refs}
:::
